Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-15 mpaper/paper.tex
Visualising the incorrect derivative

  %We begin by producing constant upper and
  %lower bounds for the elements of
  %\[
  %  \left. \frac{\partial \mathbf{K}}{\partial \lambda_i} \right|_{\lambda_i =
  %    c(\mathbf{r}, \mathbf{u})},
  %\]
  %where $\mathbf{K}$ is any covariance matrix.

@@

  % TODO: Lemma 1.1 is wrong about the % derivative. Fix (my handwritten notes
  % have the correct idea)
  Combining the bounds with the observation that
  every element of
  $\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal}$ is in
  $\mathbb{R}_2[\mathbf{u}]$ gives the required result.
\end{proof}

@@

@@

\derivatives*
\begin{proof}
  \leavevmode
  \begin{enumerate}
  \item
    \[
      \begin{split}
        \frac{\partial q(\mathbf{u})}{\partial m} &=
        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
        \\
        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
        \bm\mu] \\
        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
      \end{split}
    \]
  \item
    \begin{align}
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\exp \left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\right]\exp \left( -\frac{Q}{2} \right) \nonumber \\
                                                         &+ \frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\dB\left[\exp\left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= -\frac{1}{2} q(\mathbf{u}) \left( |\bm\Sigma|^{-3/2} \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} + \frac{\partial Q}{\partial \mathbf{B}} \right). \label{eq:2_before}
    \end{align}
    For derivatives with respect to $\bm\Sigma$, we can refer to Petersen and
    Pedersen \cite{petersen2008matrix}:
    \begin{equation} \label{eq:partial_derivatives1}
      \frac{\partial |\bm\Sigma|}{\partial \bm\Sigma} = |\bm\Sigma|\bm\Sigma^{-\intercal}, \quad \frac{\partial Q}{\partial \mathbf{B}} = -\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal},
    \end{equation}
    while we can use an online tool by Laue et
    al.\footnote{\url{http://www.matrixcalculus.org/}}
    \cite{DBLP:conf/nips/LaueMG18} for the remaining ones:
    \begin{equation} \label{eq:partial_derivatives2}
      \begin{gathered}
        \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} =
        (\adj(\mathbf{T}) + \adj(\bm\Sigma))\mathbf{B}, \\
        \dB[(\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} -
        \bm\mu)] = -(\mathbf{TUT} +
        \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1})\mathbf{B}.
      \end{gathered}
    \end{equation}
    Substituting results from \eqref{eq:partial_derivatives1} and
    \eqref{eq:partial_derivatives2} back into \eqref{eq:2_before} gives:
    \begin{alignat*}{3}
      \frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} &=
      \frac{1}{2}q(\mathbf{u})(&&\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal}), \\
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &=
      \frac{1}{2}q(\mathbf{u}) \{&&|\bm\Sigma|^{-3/2}(\adj(\mathbf{T}) +
      \adj(\bm\Sigma)) \\
      & &&+ \mathbf{TUT} + \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} \} \mathbf{B}.
    \end{alignat*}
  \item
    \begin{enumerate}
    \item Since
    \begin{align*}
      q(\mathbf{r} \mid \mathbf{u}) &= \mathcal{N}(\mathbf{r};
      \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal \Kuu^{-1}\Kru) \\
      &= \mathcal{N}(\mathbf{r}; \mathbf{Su}, \bm\Gamma),
    \end{align*}
    we have
    \[
      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
      -\frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) \dl[(\mathbf{r} -
      \mathbf{Su})^\intercal\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su}) +
      \log|\bm\Gamma|].
    \]
    The same online tool can be used to show that
    \[
      \dl \log|\bm\Gamma| = -|\bm\Gamma|^{-1}\tr(\mathbf{R}\adj(\bm\Gamma)),
    \]
    and
    \[
      \dl \bm\Gamma^{-1} = \bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}.
    \]
  \item
    If $i=0$, then
    \[
      \frac{\partial \mathbf{K}}{\partial \lambda_i} =
      \frac{1}{\lambda_i}\mathbf{K}
    \]
    by the structure of each element of $\mathbf{K}$. If $i \ne 0$, then each
    element of $\frac{\partial \mathbf{K}}{\partial \lambda_i}$ is
    \begin{align*}
      L_{j,k} &= \frac{\partial k(\mathbf{x}_j, \mathbf{x}_k)}{\partial \lambda_i} \\
      &\begin{aligned}
        = k(\mathbf{x}_j, \mathbf{x}_k) \dl \left[ \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(\mathbf{x}_j - \mathbf{x}_k)^\intercal \bm\Lambda (\mathbf{x}_j - \mathbf{x}_k) \\
        &- \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \left. \vphantom{\frac{1}{2}} \right]
      \end{aligned} \\
      &\begin{aligned}
        = k(\mathbf{x}_j, \mathbf{x}_k) \dl \left[ \vphantom{\sum_{k=1}^d} \right. &-\frac{1}{2} \sum_{l=1}^d \lambda_l (x_{j,l} - x_{k,l})^2 \\
        &- \mathbbm{1}[j \ne k]\sigma^2\sum_{l=1}^d \lambda_l \left. \vphantom{\sum_{l=1}^d} \right]
      \end{aligned} \\
      &= k(\mathbf{x}_j, \mathbf{x}_k) \left( -\frac{1}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
    \end{align*}
    \end{enumerate}
  \end{enumerate}
\end{proof}

>>>>

  Given any covariance matrix $\mathbf{K}$ and $i = 0, \dots, d$, we can easily
  establish constant upper and lower bounds on $\mathbf{K}|_{\lambda_i =
    c(\mathbf{r}, \mathbf{u})}$ by the boundedness of $c$ and continuity of the
  covariance function. Similar reasoning combined with the expressions for
  derivatives of covariance matrices in Lemma \ref{lemma:derivatives} gives
  constant upper and lower bounds on the elements of $\left.
    \frac{\mathbf{\partial K}}{\partial \lambda_i} \right|_{\lambda_i =
    c(\mathbf{r}, \mathbf{u})}$ as well. We can also use Lemma ?? to show
  constant upper and lower bounds on $\mathbf{K}^{-1}|_{\lambda_i =
    c(\mathbf{r}, \mathbf{u})}$. This is enough to prove constant upper and
  lower bounds on $\mathbf{S}$, $\bm\Gamma$, and $\mathbf{R}$, which means that
  $(\mathbf{r} -
  \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r} -
  \mathbf{Su})$ has upper and lower bounds in $\mathbb{R}_2[\mathbf{u}]$.
  % TODO: write the lemma

@@

  % TODO: Lemma 1.1 is wrong about the derivative. Fix (my handwritten notes
  % have the correct idea)
  Combining the bounds with the observation that
  every element of
  $\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal}$ is in
  $\mathbb{R}_2[\mathbf{u}]$ gives the required result.
\end{proof}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{ELBO_and_gradient_over_B}
  \caption{ELBO in blue, derivative in red. The first row is for diagonal values
  of B, the second row is for non-diagonal values.}
\end{figure*}

@@

\derivatives*
\begin{proof}
  \leavevmode
  \begin{enumerate}
  \item
    \[
      \begin{split}
        \frac{\partial q(\mathbf{u})}{\partial m} &=
        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
        \\
        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
        \bm\mu] \\
        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
      \end{split}
    \]
  \item An online tool by Laue et
    al.\footnote{\url{http://www.matrixcalculus.org/}}
    \cite{DBLP:conf/nips/LaueMG18} can be used to find both derivatives.
  \item
    \begin{enumerate}
    \item Since
    \begin{align*}
      q(\mathbf{r} \mid \mathbf{u}) &= \mathcal{N}(\mathbf{r};
      \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal \Kuu^{-1}\Kru) \\
      &= \mathcal{N}(\mathbf{r}; \mathbf{Su}, \bm\Gamma),
    \end{align*}
    we have
    \[
      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
      -\frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) \dl[(\mathbf{r} -
      \mathbf{Su})^\intercal\bm\Gamma^{-1}(\mathbf{r} - \mathbf{Su}) +
      \log|\bm\Gamma|].
    \]
    The same online tool can be used to show that
    \[
      \dl \log|\bm\Gamma| = -|\bm\Gamma|^{-1}\tr(\mathbf{R}\adj(\bm\Gamma)),
    \]
    and
    \[
      \dl \bm\Gamma^{-1} = \bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}.
    \]
  \item
    If $i=0$, then
    \[
      \frac{\partial \mathbf{K}}{\partial \lambda_i} =
      \frac{1}{\lambda_i}\mathbf{K}
    \]
    by the structure of each element of $\mathbf{K}$. If $i \ne 0$, then each
    element of $\frac{\partial \mathbf{K}}{\partial \lambda_i}$ is
    \begin{align*}
      L_{j,k} &= \frac{\partial k(\mathbf{x}_j, \mathbf{x}_k)}{\partial \lambda_i} \\
      &\begin{aligned}
        = k(\mathbf{x}_j, \mathbf{x}_k) \dl \left[ \vphantom{\frac{1}{2}} \right. &-\frac{1}{2}(\mathbf{x}_j - \mathbf{x}_k)^\intercal \bm\Lambda (\mathbf{x}_j - \mathbf{x}_k) \\
        &- \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \left. \vphantom{\frac{1}{2}} \right]
      \end{aligned} \\
      &\begin{aligned}
        = k(\mathbf{x}_j, \mathbf{x}_k) \dl \left[ \vphantom{\sum_{k=1}^d} \right. &-\frac{1}{2} \sum_{l=1}^d \lambda_l (x_{j,l} - x_{k,l})^2 \\
        &- \mathbbm{1}[j \ne k]\sigma^2\sum_{l=1}^d \lambda_l \left. \vphantom{\sum_{l=1}^d} \right]
      \end{aligned} \\
      &= k(\mathbf{x}_j, \mathbf{x}_k) \left( -\frac{1}{2}(x_{j,i} - x_{k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
    \end{align*}
    \end{enumerate}
  \end{enumerate}
\end{proof}
